Inference using context-dependent models of mutation via bounds on information propagation

### 2021-12-08 writeups/context-dependence.tex
plr edits until disco

\begin{abstract}

\end{abstract}

@@

To formalize this notion of local dependence,
first define a \textit{pattern} to mean a contiguous sequence of states,
and let $|u|$ denote the length of a pattern $u$.
Given a state $x \in \calS$,
let $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denote the pattern of $x$ of length $h$ beginning at location $i$.
% Recall we are focusing on finite sequences, and by default, we do not allow patterns to ``hang off'' the end of the sequence.
% This is natural in many situations, but in others we might want to other boundary conditions,
% modeling extant, but unobserved, neighbors.
% We will mention this when necessary.
The dynamics of our stochastic process
are determined by the set of allowed transitions and associated rates:
given two patterns $u$ and $v$ of common length $h$,
saying ``pattern $u$ changes to $v$ at rate $\mu$'' means that
\[
    \P\{ X_i^{(h)}(t+dt) = v \given X_i^{(h)}(t) = u \} = \mu dt + o(dt),
\]
regardless of the location $i$ along the sequence.
If more than one pattern would match to cause the same change, their rates add.
Such a rule is defined by its ``transition triple'' $(\mu,u,v)$,
written as ``$u \to v$ at rate $\mu$''.
A model of context-dependent mutation
is defined by a set of all transition triples,
which we denote $\calT$.

@@

  \noindent
  This only has one parameter.
  If we allow particles to exit from the right, and new particles to enter from the left,
  then from observing only starting and ending configurations
  we cannot tell with certainty which particles have moved where,
  so it is not obvious how to estimate the speed.
  However, summing over possible particle movements would allow us to
  compute the likelihood of a given configuration change,
  and so estimate the speed by maximum likelihood
  given an initial sequence, a final sequence and an elapsed time.

@@

Concretely, let $\calP = \{(p_i, e_i)\}$ be a set of (pattern, energy difference) pairs,
and for any sequence $x$ define
$E(x) = \sum_i e_i\, n(x,p_i)$,
where $n(x,p_i)$ is the number of times the pattern $p_i$ occurs in $x$.
These will affect the rates through a nonnegative function $\phi$:
we declare that if the rate at which $x$ changes to $y$ as computed
from transition triples is $\mu$,
then the actual rate is
\[
    \mu \, \phi\left(E(y) - E(x)\right) .
\]
If $\phi(e)$ is not greater than 1, one way to think about this is that
the transition triples give ``proposed changes'',
but these changes only take effect with probability $\phi(E(y) - E(x))$.
We have described this idea in terms natural for a physics model such as an Ising model in which high energy configurations are disfavored, however it is also suitable for a population genetics model in which $\phi$ describes the probability of a mutant allele given a certain fitness change from the parent.

\begin{example}[Genomic GC content]
    A genome sequence can be written using A, C, G, and T;
    in the most general model of independent mutation across sites, each of the 12 possible transitions occurs at its own rate.
    Furthermore, it is a well-known observation that in many species,
    adjacent, methylated CG dinucleotides (``CpG sites'') have a much higher mutation rate to TG and CA
    than either single nucleotide change under the independent model.
    Embellishing the single-nucleotide model with this additional rate results in the model defined by
    \begin{center}
      \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }cc}
        $x$  &  $y$  &  $m_{xy}$ & for $x, y \in \{\nA,\nC,\nG,\nT\}$ and $x \neq y$ \\
        \nC\nG   &  \nT\nG   &  $\gamma$ & \\
        \nC\nG   &  \nC\nA   &  $\gamma$ &
      \end{tabular},
    \end{center}
    where $\gamma$ is the additional CpG rate above the base mutation rate.
    Note that for a given pair of sequences,
    a change $\nA\nC\nG \to \nA\nT\nG$ could have occurred by a $\nC \to \nT$ mutation
    via the independent-sites model,
    or by a $\nC\nG \to \nT\nG$ mutation;
    in such cases
    the total instantaneous rate at which ACG changes to ATG is the sum of the rates
    (in this example, $\gamma + m_{\nC\nT}$).
    % Although this means there are many possible parameterizations,
    % note that the model is identifiable, not overparameterized.

    Counteracting this trend towards increased G/C is GC-biased gene conversion \citep{glemin2015quantification},
    which is effectively a selective pressure against G or C bases.
    We model the evolution of a \emph{single} sequence, not a population of sequences,
    imagining this sequence to be the consensus sequence of the population.
    A new mutation with an effect $s$ on fitness that occurs
    in a population of effective size $N_e$ becomes ubiquitous, rather than dying out,
    with probability approximately $(1-\exp(-2 s))/(1-\exp(-2 s N_e))$ \citep{kimura1962probability}.
    If the mutation occurs at rate $\mu$ per individual,
    the total rate it occurs at in the population is $\mu N_e$.
    GC-biased gene conversion effectively means that sequences with more G/Cs are selected against.
    This is incorporated as follows: set the energy of pattern of pattern \nC{} or \nG{} to be $s$,
    so that $E(y)-E(x)$ is the net change in number of G's and C's multiplied by $s$.
    Also, define $\phi(e)$ to be the expected rate of fixations of ``selected'' mutations
    with advantage $e$:
    \begin{align*}
        \phi(e) = N_e (1-\exp(-2e))/(1-\exp(-2eN)).
    \end{align*}
    With these definitions,
    if a given change that occurs with rate $\mu$ would change a sequence $x$ to $y$,
    then $\mu \, \phi(E(y)-E(x))$ is the rate at which the mutation
    appears and successfully takes over in the population.

@@

@@

First, it will be useful for later proofs to have an explicit construction of the process,
also used for simulation,
using a variant of the ``jump chain representation'', also known as the ``Gillespie algorithm''
or ``uniformization'' \citep{hobolth2009simulation}.
Briefly, this  works by first sampling a homogeneous Poisson process of times and locations of possible changes,
at an appropriate rate $\mu^*$ per site,
and then resolving each possible change in temporal order.
(Note that these ``possible changes'' are different than the ``proposed changes'' described above
in terms of the energy model.)
The rate $\mu^*$ should be the maximum ``local'' rate at which transitions occur, across sites and transition outcome states:
\[
    \mu^*
    % = \max \left\{ \sum_{y \in \calS^L} \sum_{(\mu, u, v) \in J(i,x,y)} \mu \qquad \st x \in \calS^L, \quad 1 \le i \le L \right\} .
    = \max_x \sum_{y \neq x} G(L)_{x,y} .
\]
Note that we do not actually need to construct $G(L)$ to find the maximum rate.
Now sample possible changes:
in a sequence of length $L$, over a time period $t$,
first draw the number of possible changes, $N$,
from a Poisson distribution with mean $\mu^* t L$,
and then choose the times $s_k$ and locations $i_k$
at which possible changes occur uniformly
over $[0, t]$ and $\{1, 2, \ldots, L\}$ respectively.
Order the possible change times, so that
\[
    0 < s_1 < \cdots < s_N < t .
\]
(Note that this is equivalent to saying that possible
changes occur at rate $\mu^*$ independently at each of the $L$ sites.)
Now, suppose we have determined the state just before time $s_k$ to be $X(s_k) = x$.
Then, a new state is then chosen at position $i_k$
with probability proportional to the corresponding rate, i.e.,
out of the possible transitions $\calT$, transition triple $(\mu(j), u(j), v(j))$ is chosen
with probability
\[
q_j = \begin{cases}
    \mu(j)/\mu^* \qquad & \text{if } x_i^{(|u(j)|)} = u(j)  \\
    0 \qquad & \text{otherwise.}
\end{cases}
\]
The remaining probability $q_0 = 1-\sum_{j=1}^{|\calT|} q_j$
gives the probability that the state remains the same.
(By construction of $\mu_*$, $q_0 \ge 0$.)
If triple $(\mu(j), u(j), v(j))$ is chosen, then
the substring of $x$ matching $u(j)$ that begins at position $i_k$
is replaced with $v(j)$ to produce $X(s_k)$.

@@

The key insight we use is that by including neighboring sequence context in the \emph{initial} sequence only,
we can compute \emph{something} correctly,
and that will be sufficient to compute the full likelihood
to an arbitrarily good approximation.
%EM do we really want x and y here, which I've been thinking of as being the full sequence?
%EM Note we'll need to change a and b below.
To do this, let $p_{n,\ell,r}(x,y)$ for $\ell + r < n$ denote
the probability that an initial subsequence $x$ of length $n$, after time $t$,
is found to match a smaller subsequence $y$ of length $n-\ell-r$, offset by $\ell$ sites.
This central, smaller subsequence is called the ``base'' of the T-mer.
We refer to such patterns as \emph{T-mers}, as depicted in Figure~\ref{fig:Tmers}.
We emphasize that the overhang of the T-mers is typically bigger than the
context-sensitivity width of the underlying model (Figure~\ref{fig:overview}).

@@

What we actually want is the \emph{full likelihood} $\like(y|x)$,
the probability that a given sequence $x$ evolves into another, $y$, over a given period of time.
This can be computed as the product of the probability of each site in $y$
given $x$ and all sites in $y$ further to the left.
Defining $x_{<i}=(x_1, \ldots, x_{i-1})$ to be the subsequence to the left of $i$, this is:
\begin{align*}
    \like(y \mid x) &= \prod_{i=1}^L \P\{ Y_i = y_i \given Y_{<i} = y_{<i} \ \text{and}\; X=x \} .
\end{align*}
Each term in this product is the probability of $Y_i$, given all sites in $X$
and all sites in $Y$ to the left of $i$.
We can compute an approximation to this probability as in \eqref{eqn:window_approx}
by discarding those sites from $X$ and $Y$ that are approximately independent of $Y_i$.
Which are these?

These approximately independent sites are exactly those that are too far away from a given site to impact its evolution.
Recall that the last base of $x_i^{(h)}$ is $x_{i+h-1}$.
So, $x_{i-3\ell}^{(4\ell+1)}$ runs from $i-3\ell$ to $i + \ell$, inclusive, while $y_{i-2\ell}^{(2\ell+1)}$ runs from $i-2\ell$ to $i$, inclusive.
If $\ell$ is long enough that approximation~\eqref{eqn:window_approx} holds, then:
\begin{align} \label{eqn:full_approx}
    \begin{split}
    \P\{ Y_i = y_i \given Y_{<i} \ \text{and}\; X=x \}
    &\approx
        \P\{ Y_i = y_i \given Y_{i-2\ell}^{(2\ell+1)} \;\text{and}\; X_{i-3\ell}^{(4\ell+1)} \}  \\
    &=
        \frac{
            p_{2\ell+1,\ell+1,\ell}(x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell+1)})
        }{
            p_{2\ell+1,\ell,\ell+1}(x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell)})
        }.
    \end{split}
\end{align}

We can then simply express our likelihood as
\[
    \like(y \mid x) = \frac{ A_{2\ell+1, \ell+1, \ell}(x,y) }{ A_{2\ell+1, \ell, \ell+1}(x,y) }.
\]
% Define $n_\ell(x,y;a,b)$ to be the number of times pattern $a$ is seen in sequence $x$
% matched to pattern $b$ in $y$ at a position offset by $\ell$.
% Then
% \begin{align}
%     \log \like(y|x)
%     &\approx
%         \sum_{a \in \calS^n} \sum_{b \in \calS^{n-2\ell}} n_\ell(x,y;a,b) \log p_{n,\ell,\ell}(a,b)
%         -
%         \sum_{c \in \calS^n} \sum_{d \in \calS^{n-2\ell-1}} n_\ell(x,y;c,d) \log p_{n,\ell,\ell+1}(c,d)
% \end{align}

@@

See Appendix~\ref{ss:approx_pf} for a more rigorous statement and proof,
and the next section for some intuition.

@@

For instance, in the CpG model the window width is $w=1$
because the process only depends on neighboring sites;
and if the sequence has evolved for time $t$ then we expect no more than a Poisson($\lambda t$) number of changes at each site.
If $\lambda t = 0.75$ (three-quarters of the sequence will have changed!),
then taking $\ell = r = 3$ allows us to compute Tmer probabilities to within an error of 0.007,
since this is the probability that a Poisson with mean 0.75 will be larger than 3.
If $\lambda t = 0.25$ we would do even better with $\ell = r = 2$.

To obtain the full likelihood, in \eqref{eqn:full_approx}, in addition to an overhang of $\ell$
we needed the base of the T-mer to be of length $2\ell+1$.
This is because we have also conditioned on the final sequence to the left of the focal site, $Y_{<i}$,
and although the process at $i$ only depends on events within $\pm \ell$,
these events may affect other sites in $Y$ going back a distance $2\ell$.

@@

For inference, we need to compute the $|\calS|^{n} \times |\calS|^m$ matrix $F$ whose $(a,b)^\text{th}$ entry is $p_{n,\ell,r}(t,a,b)$,
for various $t$.
This matrix is a projection of the $|\calS|^{n} \times |\calS|^{n}$ matrix whose $(a,c)^\text{th}$ entry is
\begin{align}
    p_{n}(a,c) = \left( e^{t G(n)} \right)_{a,c} ,
\end{align}
where $G(n)$ is the sparse $|\calS|^{n} \times |\calS|^{n}$ matrix defined in \eqref{eqn:G_defn}.
The ``projection'' we need just marginalizes over long (i.e.\ length $n$) patterns $c$ that match the shorter (i.e.\ length $m$) pattern $b$
as in \eqref{eq:pmMarg}:
if we define the $n \times m$ matrix $U$ so that $U_{cb}=1$ if $c_\ell^m=b$ and $U_{cb}=0$ otherwise,
then
\begin{align} \label{eqn:Tmer_trans}
    p_{n,\ell,r}(t,a,b) = \sum_{c \in \calS^{n}} \left( e^{t G(n)} \right)_{ac} U_{cb} ,
\end{align}
so in fact we don't need the entire matrix $e^{t G(n)}$,
just the product of this matrix with each of the columns of $U$.
Modern techniques in sparse matrix computation (e.g., Krylov methods) provide efficient ways to do this.

@@

Although we haven't used such tools here, taking the derivative with respect to $t$ can also be performed using sparse matrix operations.

\paragraph{Sparsity and updating $G$}
We can also take advantage of the sparsity of $G$ to perform efficient computation of the likelihood
under many sets of parameters.
Let's assume that the T-mers only allow single-position changes.
First note that in this case $G(n)$ has
$(1+m(|\calS|-1)) |\calS|^{n}$ nonzero entries, since each of the $|\calS|^{n}$ can change in $m$ places.
This will determine how the computation scales with $m$, $n$, and $|\calS|$,
once we precompute a number of things.
Let $g = (g_1, \ldots, g_d)$ be the nonzero entries of $G$, in some fixed ordering;
sparse matrix representations of $G$ store only $g$ along with information about the rows and columns these are found in.
%EM Here we are using lower j, rather than our "new" notation setup, but it doesn't seem like a big deal. This is a local note.
Each $g_i$ is a linear combination of mutation rates $\mu_j$,
multiplied by a function ($\phi$) of a linear combination of energy coefficents $e_j$,
say $g_i = \sum_j A_{ij} \mu_j \phi(\sum_k B_{ik} e_k)$,
so by precomputing the matrices $A$ and $B$ we can update $G$ with new parameter values
using only two matrix multiplications and evaluation of $\phi()$ over a vector.
This remains efficient to perform in interpreted languages such as R,
since each step is carried out by lower-level compiled code (e.g., optimized linear algebra libraries).
% This is true if the boundary conditions are circular, mean-value, or neither.

@@

\begin{figure}
    \begin{center}
    \includegraphics{peeling-schematic}
    \end{center}
    \caption{
        A depiction of the four steps in computing the likelihood of finding particular ``short'' subsequence at $Y$ and $Z$
        given the ``long'' subsequence at $X$: see text for details.
        \label{fig:peeling}
    }
\end{figure}

@@

%%%%%%%%%%%
\subsection{Hominid data}

@@

In exploratory application to real data, how does one decide when to stop adading mutational motifs?

@@

(c) repeat until no residuals are statisically significant.

@@

@@

The usage of fast sparse matrix methods described above make likelihood computation fast enough
to use a Metropolis-Hasting-based Markov chain Monte Carlo method
to sample from the posterior distribution on parameters.
To test this,
we evolved $10^6$ sites from the Ising model (described above) with $\lambda=0.5$, $\beta=1.0$, and $\gamma=0.5$
for 1 unit of time,
resulting in around 460,000 possible mutations
and differences at around 186,000 sites between initial and final sequence.
We then placed an Exponential prior with mean 1 on both transitions $+ \to -$ and $- \to +$
(without constraining these to be equal)
and half-Gaussian priors with mean 0 and scale 3 on both $\beta$ and $\gamma$,
counted $(5,3,1)$ T-mers,
and used the mcmc package \citep{geyer2017mcmc} to run a random walk sampler for 10,000 steps
(judged sufficient by examination of diagnostic plots).
We repeated this procedure separately on 60 independently simulated sequences,
and show the resulting credible intervals in Figure~\ref{fig:ising_coverage}.

@@

We also applied the method to changes between the human and chimpanzee reference sequences.

>>>>

\begin{abstract}
    Although the rates at which positions in the genome mutate
    are known to depend not only on the nucleotide to be mutated
    but also on neighboring nucleotides,
    it remains challenging to do phylogenetic inference
    using models of context-dependent mutation.
    In these models,
    the effects of one mutation may in principle propagate
    to faraway locations,
    making it difficult to compute exactly likelihoods.
    This paper shows how to use bounds on the propagation of dependency
    to compute likelihoods of mutation of a given segment of genome
    by marginalizing over sufficiently long flanking sequence.
    This can be used for maximum likelihood or Bayesian inference.
    Protocols examining residuals and iterative model refinement
    are also discussed.
    Tools for efficiently working with these models
    are provided in an R package,
    that could be used in other applications.
    The method is used to examine context dependence of mutations
    since the common ancestor of humans and chimpanzee.
\end{abstract}

@@

To formalize this notion of local dependence,
first define a \textit{pattern} to mean a contiguous sequence of states,
and let $|u|$ denote the length of a pattern $u$.
Given a sequence $x \in \calS^L$,
let $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ denote the pattern of $x$ of length $h$ beginning at location $i$.
% Recall we are focusing on finite sequences, and by default, we do not allow patterns to ``hang off'' the end of the sequence.
% This is natural in many situations, but in others we might want to other boundary conditions,
% modeling extant, but unobserved, neighbors.
% We will mention this when necessary.
The dynamics of our stochastic process
are determined by the set of allowed transitions and associated rates:
given two patterns $u$ and $v$ of common length $h$,
saying ``pattern $u$ changes to $v$ at rate $\mu$'' means that
\[
    \P\{ X_i^{(h)}(t+dt) = v \given X_i^{(h)}(t) = u \} = \mu dt + o(dt),
\]
regardless of the location $i$ along the sequence.
If more than one pattern would match to cause the same change, their rates add.
Such a rule is defined by its ``transition triple'' $(\mu,u,v)$,
written as ``$u \to v$ at rate $\mu$''.
A model of context-dependent mutation
is defined by a set of all transition triples,
which we denote $\calT$.

@@

  \noindent
  This only has one parameter.
  If we allow particles to exit from the right and new particles to enter from the left
  (e.g., by adding special symbols for the start and end of the sequence),
  then from observing only starting and ending configurations
  we cannot tell with certainty which particles have moved where,
  so it is not obvious how to estimate the speed.
  However, summing over possible particle movements would allow us to
  compute the likelihood of a given configuration change,
  and so estimate the speed by maximum likelihood
  given an initial sequence, a final sequence and an elapsed time.

@@

Concretely, let $\calP = \{(p_i, e_i)\}$ be a set of (pattern, energy difference) pairs,
and for any sequence $x$ define
$E(x) = \sum_i e_i\, n(x,p_i)$,
where $n(x,p_i)$ is the number of times the pattern $p_i$ occurs in $x$.
These will affect the rates through a nonnegative function $\phi$:
we declare that if the rate at which $x$ changes to $y$ as computed
from transition triples is $\mu$,
then the actual rate is
\[
    \mu \, \phi\left(E(y) - E(x)\right) .
\]
If $\phi(e)$ is not greater than 1, one way to think about this is that
the transition triples give ``proposed changes'',
but these changes only take effect with probability $\phi(E(y) - E(x))$.
We have described this idea in terms natural for a physics model
in which high energy configurations are disfavored,
however it is also suitable for a population genetics model
in which $\phi$ describes the probability of fixation of a mutant allele
given a certain fitness change from the parent.

\begin{example}[Genomic GC content]
    A genome sequence can be written using A, C, G, and T;
    in the most general model of independent mutation across sites, each of the 12 possible transitions occurs at its own rate.
    Furthermore, in many species
    adjacent, methylated CG dinucleotides (``CpG sites'') have a much higher mutation rate to TG and CA
    than either single nucleotide change under the independent model.
    Embellishing the single-nucleotide model with this additional rate results in the model defined by
    \begin{center}
      \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }cc}
        $x$  &  $y$  &  $m_{xy}$ & for $x, y \in \{\nA,\nC,\nG,\nT\}$ and $x \neq y$ \\
        \nC\nG   &  \nT\nG   &  $\gamma$ & \\
        \nC\nG   &  \nC\nA   &  $\gamma$ &
      \end{tabular},
    \end{center}
    where $\gamma$ is the additional CpG rate above the base mutation rate.
    Note that for a given pair of sequences,
    a change $\nA\nC\nG \to \nA\nT\nG$ could have occurred by a $\nC \to \nT$ mutation
    via the independent-sites model,
    or by a $\nC\nG \to \nT\nG$ mutation;
    in such cases
    the total instantaneous rate at which ACG changes to ATG is the sum of the rates
    (in this example, $\gamma + m_{\nC\nT}$).
    % Although this means there are many possible parameterizations,
    % note that the model is identifiable, not overparameterized.

    Counteracting this trend towards increased G/C is GC-biased gene conversion \citep{glemin2015quantification},
    which acts effectively as a selective pressure against G or C bases.
    We model the evolution of a \emph{single} sequence, not a population of sequences,
    imagining this sequence to be the consensus sequence of the population.
    A new mutation with an effect $s$ on fitness that occurs
    in a population of effective size $N_e$ becomes ubiquitous, rather than dying out,
    with probability approximately $(1-\exp(-2 s))/(1-\exp(-2 s N_e))$ \citep{kimura1962probability}.
    If the mutation occurs at rate $\mu$ per individual,
    the total rate it occurs at in the population is $\mu N_e$.
    GC-biased gene conversion effectively means that sequences with more G/Cs are selected against.
    This is incorporated as follows: set the energy of patterns \nC{} and \nG{} to be $s$,
    so that $E(y)-E(x)$ is the net change in number of G's and C's multiplied by $s$.
    Also, define $\phi(e)$ to be the expected rate of fixations of ``selected'' mutations
    with advantage $e$ relative to the per-individual, per-site mutation rate:
    \begin{align*}
        \phi(e) = N_e (1-\exp(-2e))/(1-\exp(-2eN)).
    \end{align*}
    With these definitions,
    if a given change that occurs with rate $\mu$ would change a sequence $x$ to $y$,
    then $\mu \, \phi(E(y)-E(x))$ is the rate at which the mutation
    appears and successfully takes over in the population.

@@

We close with a final example from statistical physics, to be used later:

@@

First, it will be useful for later proofs to have an explicit construction of the process,
also used for simulation,
using a variant of the ``jump chain representation'', also known as the ``Gillespie algorithm''
or ``uniformization'' \citep{hobolth2009simulation}.
Briefly, this  works by first sampling a homogeneous Poisson process of times and locations of possible changes,
at an appropriate rate $\mu^*$ per site,
and then resolving each possible change in temporal order.
(Note that these ``possible changes'' are different than the ``proposed changes'' described above
in terms of the energy model.)
The rate $\mu^*$ should be the maximum ``local'' rate at which transitions occur, across sites and transition outcome states:
\[
    \mu^*
    % = \max \left\{ \sum_{y \in \calS^L} \sum_{(\mu, u, v) \in J(i,x,y)} \mu \qquad \st x \in \calS^L, \quad 1 \le i \le L \right\} .
    = \max_x \sum_{y \neq x} G(L)_{x,y} .
\]
Note that we do not actually need to construct $G(L)$ to find the maximum rate.
Now sample possible changes:
in a sequence of length $L$, over a time period $t$,
first draw the number of possible changes, $N$,
from a Poisson distribution with mean $\mu^* t L$,
and then choose the times $s_k$ and locations $i_k$
at which possible changes occur uniformly
over $[0, t]$ and $\{1, 2, \ldots, L\}$ respectively.
Order the possible change times, so that
\[
    0 < s_1 < \cdots < s_N < t .
\]
(Note that this is equivalent to saying that possible
changes occur at rate $\mu^*$ independently at each of the $L$ sites.)
Now, suppose we have determined the state just before time $s_k$ to be $X(s_{k-1}) = x$.
Then, the new state $X(s_k)$ is chosen by applying a mutation at position $i_k$
with probability proportional to the corresponding rate, i.e.,
out of the possible transitions $\calT$, transition triple $(\mu(j), u(j), v(j))$ is chosen
with probability
\[
q_j = \begin{cases}
    \mu(j)/\mu^* \qquad & \text{if } x_i^{(|u(j)|)} = u(j)  \\
    0 \qquad & \text{otherwise.}
\end{cases}
\]
The remaining probability $q_0 = 1-\sum_{j=1}^{|\calT|} q_j$
gives the probability that the state remains the same.
(By construction of $\mu_*$, $q_0 \ge 0$.)
If triple $(\mu(j), u(j), v(j))$ is chosen, then
the substring of $x$ matching $u(j)$ that begins at position $i_k$
is replaced with $v(j)$ to produce $X(s_k)$.

@@

The key insight we use is that by including neighboring sequence context in the \emph{initial} sequence only,
we can compute \emph{something} correctly,
and that will be sufficient to compute the full likelihood,
at least arbitrarily.
%EM do we really want x and y here, which I've been thinking of as being the full sequence?
%EM Note we'll need to change a and b below.
To do this, let $p_{n,\ell,r}(x,y)$ for $\ell + r < n$ denote
the probability that an initial subsequence $x$ of length $n$, after time $t$,
is found to match a smaller subsequence $y$ of length $n-\ell-r$, offset by $\ell$ sites.
This central, smaller subsequence is called the ``base'' of the T-mer.
We refer to such patterns as \emph{T-mers}, as depicted in Figure~\ref{fig:Tmers}.
We emphasize that the overhang of the T-mers is typically bigger than the
context-sensitivity width of the underlying model (Figure~\ref{fig:overview}).

@@

What we actually want is the \emph{full likelihood} $\like(y|x)$,
the probability that a given sequence $x$ evolves into another, $y$, over a given period of time.
This can be computed as the product of the probability of each site in $y$
given $x$ and all sites in $y$ further to the left.
Defining $x_{<i}=(x_1, \ldots, x_{i-1})$ to be the subsequence to the left of $i$, this is:
\begin{align*}
    \like(y \mid x) &= \prod_{i=1}^L \P\{ Y_i = y_i \given Y_{<i} = y_{<i} \ \text{and}\; X=x \} .
\end{align*}
Each term in this product is the probability of $Y_i$, given all sites in $X$
and all sites in $Y$ to the left of $i$.
It seems reasonable that, similar to equation \eqref{eqn:window_approx} above,
the process at position $i$ should be approximately independent
of distant positions, given the neighboring positions.
Suppose that $\ell$ is long enough that approximation~\eqref{eqn:window_approx} holds
with $\ell = r$.
Then we know that site $i$ approximately depends only on mutations within distance $\ell$.
Furthermore, the sites up to distance $2 \ell$ to the left of $i$
also depend on these mutations.
We can now use \eqref{eqn:window_approx} to restrict the dependency on $x$.
Recall that the last base of $x_i^{(h)}$ is $x_{i+h-1}$.
So, $x_{i-3\ell}^{(4\ell+1)}$ runs from $i-3\ell$ to $i + \ell$, inclusive, while $y_{i-2\ell}^{(2\ell+1)}$ runs from $i-2\ell$ to $i$, inclusive.
This suggests that the following approximation should be good:
\begin{align} \label{eqn:full_approx}
    \begin{split}
    \P\{ Y_i = y_i \given Y_{<i} \ \text{and}\; X=x \}
    &\approx
        \P\{ Y_i = y_i \given Y_{i-2\ell}^{(2\ell+1)} \;\text{and}\; X_{i-3\ell}^{(4\ell+1)} \}  \\
    &=
        \frac{
            p_{2\ell+1,\ell+1,\ell}(x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell+1)})
        }{
            p_{2\ell+1,\ell,\ell+1}(x_{i-3\ell}^{(4\ell+1)}, y_{i-2\ell}^{(2\ell)})
        }.
    \end{split}
\end{align}

Since the full likelihood is a product across sites,
we can then simply express an approximate full likelihood as
a ratio of two composite likelihoods:
\[
    \like(y \mid x) \approx \frac{ A_{2\ell+1, \ell+1, \ell}(x,y) }{ A_{2\ell+1, \ell, \ell+1}(x,y) }.
\]
% Define $n_\ell(x,y;a,b)$ to be the number of times pattern $a$ is seen in sequence $x$
% matched to pattern $b$ in $y$ at a position offset by $\ell$.
% Then
% \begin{align}
%     \log \like(y|x)
%     &\approx
%         \sum_{a \in \calS^n} \sum_{b \in \calS^{n-2\ell}} n_\ell(x,y;a,b) \log p_{n,\ell,\ell}(a,b)
%         -
%         \sum_{c \in \calS^n} \sum_{d \in \calS^{n-2\ell-1}} n_\ell(x,y;c,d) \log p_{n,\ell,\ell+1}(c,d)
% \end{align}

@@

See the next section for more discussion of the approximations.

@@

For instance, in the CpG model the window width is $w=1$
because the process only depends on neighboring sites;
and if the sequence has evolved for time $t$ then we expect no more than a Poisson($\lambda t$) number of changes at each site.
If $\lambda t = 0.75$ (three-quarters of the sequence will have changed!),
then taking $\ell = r = 3$ allows us to compute T-mer probabilities to within an error of 0.007,
since this is the probability that a Poisson with mean 0.75 will be larger than 3.
If $\lambda t = 0.25$ we would do even better with $\ell = r = 2$.

Appendix~\ref{ss:approx_pf} makes this argument more precise;
however, the argument for the full likelihood is slipperier,
and we can only provide the rough argument above.
To see one difficulty, consider the TASEP model,
with $x_1 = y_n = 1$, and all remaining positions equal to 0.
The full probability of this pair of sequences is very small,
while the conditional probability of $Y_n=1$ given $X=10\cdots0$ and $Y_{<n} = 0\cdots0$
is large since the 1 that initially began in the first position
had to go somewhere.
However, if we do not include the first position in the conditioning,
this conditional probability is very different.
% More generally, it is harder because conditioning on the final sequence $y$
% in effect allows dependencies to propagate back in time as well as forward.
Fortunately, such situations are rare,
and the TASEP model provides the extreme case of dependency propagation.
In practice, we recommend that the likelihood is computed with equation \eqref{eqn:full_approx}
using successively larger values of $\ell$
to check for convergence.

@@

For inference, we need to compute the $|\calS|^{n} \times |\calS|^m$ matrix $F$ whose $(a,b)^\text{th}$ entry is $p_{n,\ell,r}(t,a,b)$,
for various $t$ (and $m = n - \ell - r$).
This matrix is a projection of the $|\calS|^{n} \times |\calS|^{n}$ matrix whose $(a,c)^\text{th}$ entry is
\begin{align}
    p_{n}(a,c) = \left( e^{t G(n)} \right)_{a,c} ,
\end{align}
where $G(n)$ is the sparse $|\calS|^{n} \times |\calS|^{n}$ matrix defined in \eqref{eqn:G_defn}.
The ``projection'' we need just marginalizes over long (i.e., length $n$) patterns $c$ that match the shorter (i.e., length $m$) pattern $b$
as in \eqref{eq:pmMarg}:
if we define the $n \times m$ matrix $U$ so that $U_{cb}=1$ if $c_\ell^{(m)}=b$ and $U_{cb}=0$ otherwise,
then
\begin{align} \label{eqn:Tmer_trans}
    p_{n,\ell,r}(t,a,b) = \sum_{c \in \calS^{n}} \left( e^{t G(n)} \right)_{ac} U_{cb} ,
\end{align}
so in fact we don't need the entire matrix $e^{t G(n)}$,
just the product of this matrix with each of the columns of $U$.
Modern techniques in sparse matrix computation (e.g., Krylov methods) provide efficient ways to do this.

@@

% Although we haven't used such tools here, taking the derivative with respect to $t$ can also be performed using sparse matrix operations.

\paragraph{Sparsity and updating $G$}
We can also take advantage of the sparsity of $G$ to perform efficient computation of the likelihood
under many sets of parameters.
Let's assume that the T-mers only allow single-position changes.
First note that in this case $G(n)$ has
$(1+n(|\calS|-1)) |\calS|^{n}$ nonzero entries,
since each of the $|\calS|^{n}$ sequences can change in $n$ places.
This will determine how the computation scales with $n$ and $|\calS|$,
once we precompute a number of things.
Let $g = (g_1, \ldots, g_d)$ be the nonzero entries of $G(n)$, in some fixed ordering;
sparse matrix representations of $G(n)$ store only $g$ along with information about the rows and columns these are found in.
Each $g_i$ is a linear combination of mutation rates $\mu_j$,
multiplied by a function ($\phi$) of a linear combination of energy coefficents $e_j$,
say $g_i = \sum_j A_{ij} \mu_j \phi(\sum_k B_{ik} e_k)$,
so by precomputing the matrices $A$ and $B$ we can update $G(n)$ with new parameter values
using only two matrix multiplications and evaluation of $\phi()$ over a vector.
This remains efficient to perform in interpreted languages such as R,
since each step is carried out by lower-level compiled code (e.g., optimized linear algebra libraries).
% This is true if the boundary conditions are circular, mean-value, or neither.

@@

\begin{figure}
    \begin{center}
    \includegraphics{peeling-schematic}
    \end{center}
    \caption{
        A depiction of the four steps in computing the likelihood
        of finding particular ``short'' subsequences at $Y$ and $Z$
        given the ``long'' subsequence at $X$: see text for details.
        \label{fig:peeling}
    }
\end{figure}

@@

%%%%%%%%%%%
\subsection{Hominid data}
\label{hominid_section}

@@

In exploratory application to real data, how does one decide when to stop adding mutational motifs?

@@

(c) repeat until no residuals are statistically significant.

@@

These were indeed the motifs used when simulating the data.

@@

The usage of fast sparse matrix methods described above make likelihood computation fast enough
to use a Metropolis-Hasting-based Markov chain Monte Carlo method
to sample from the posterior distribution on parameters.
To test this,
we evolved $10^6$ sites from the Ising model (described above) with $\lambda=0.5$, $\beta=1.0$, and $\gamma=0.5$
for 1 unit of time,
resulting in around 460,000 possible mutations
and differences at around 186,000 sites between initial and final sequence.
We then placed an Exponential prior with mean 1 on both transitions $+ \to -$ and $- \to +$
(without constraining these to be equal)
and half-Gaussian priors with mean 0 and scale 3 on both $\beta$ and $\gamma$,
counted $(5,3,1)$ T-mers,
and used the mcmc package \citep{geyer2017mcmc} to run a random walk sampler for 10,000 steps
(judged sufficient by examination of diagnostic plots).
We repeated this procedure separately on 61 independently simulated sequences,
and show the resulting credible intervals in Figure~\ref{fig:ising_coverage}.

@@

We also applied the method to changes between the human and chimpanzee reference sequences,
using the model described above.
